\papertitle

### 2014-03-20 final-report/paper.tex
refactoring intro

We propose \system, a visualization tool that helps database developers/users
explore and understand query execution and data movement in a distributed
database management system (DDBMS). \system provides quick insight into common
problems such as data skew or performance bottleneck by leveraging
visualization techniques to present: (1) data flow between query segments and
between workers, (2) query execution and operator dependencies, (3) cluster
utilization, (4) network utilization.

@@

Database query performance is often unpredictable, especially when the DBMS is
distributed. This requires programmers or database administrators to spend
significant amounts of time and effort isolating and debugging the root cause
of performance problems. These may include skewed data partitioning or run-time
issues, such as resource contention or adverse network conditions.

Our goal is to build a visualization tool designed to help a database administrator understand the data flow and run-time
performance during query execution in a Distributed Database Management System (DDBMS). Using our tool, the
database programmer or administrator could quickly discover run-time bottlenecks as well as problematic data skew issues,
saving them significant guesswork and programming effort to identify the appropriate logs to analyze. Our visualization tool,
\system, uses profiling logs from \emph{Myria} \cite{myria} as the prototype DDMS. Myria is a distributed
database platform that provides \emph{big data management and analytics as a service} primarily for scientific applications.
Myria takes queries in high-level query languages, such as SQL, translating them into physical query plans and finally executing
them in parallel. Currently, Myria does not offer a visualization tool designed to show the physical query plan, job execution and network
communication between distributed workers. \system allows the programmer/database administrator to visualize the following aspects
of query execution.
\todo{it's not a dag, as we noticed}

\begin{itemize}

\item An interactive visualization of the physical query plan. The physical
query plan is composed of a DAG of query \emph{fragments}. A query
fragment is a subtree of the query execution plan that is separated from the
rest of the plan by a network boundary. This means that the operators inside a
query fragment can be executed by a single thread. Each fragment is scheduled
to run on multiple workers, which means each worker thread executes the same set of
operators inside the fragment, but on different data. Each DAG link represents
communication between workers, possibly over the network.  \system displays
this DAG in an interactive fashion with clickable elements for fragment nodes
and communication links.  This DAG also represents a high-level data flow in
the DDBMS.

@@

The role of the back-end is produce event logs that record when operators are called, when the call returns and when data is sent to another worker. These logs are then used to create the visualizations in \system's web UI as described in Section~\ref{sec:front}.

There are two possible approaches that we explored to collect event logs. The first and simplest approach was to write the logs on files directly to disk. The files would then get hauled back on one node using remote copy. The logs would then be parsed, and the data transformed into the desired format.

We switched to a more scalable data collection approach approach as shown in Figure~\ref{fig:arch} where logs get written directly into the worker's database as tuples in a relation. When the front-end client requests data, the master executes a query that collects the data and streams the results directly to the requesting client. We found this data collection approach to be more scalable, reliable and faster. More particularly, we are able to write queries to aggregate, filter and transform the collected data.

For example, when querying how many workers are working on a certain fragment at a given time to produce the visualization described in Section~\ref{sec:fragment}, we used a query to select only events in the root operators (the operators that have no parent in the same fragment) for each fragment and a custom map function that carries state. The state is an integer that is incremented when the operator is called on a worker and decremented when the operator returns.

@@

\subsection{Front-end}
\label{sec:front}

The Myria web front-end server is written in Python and runs on Google App Engine\footnote{\url{https://developers.google.com/appengine/}}. \system's user-interface (UI) is embedded into Myria's web front-end. We build \system's UI using D3\cite{2011-d3} to deliver an interactive visualization experience to the user. D3 is a JavaScript framework for data visualization on the web.

\system's front-end gets data from the Myria server as JSON and CSV files. JSON is used to describe the physical queryplan. CSV files contain logs on query execution at the fragment and operator level as well as data exchange between workers at different stages of the query execution.

@@

\subsubsection{Overview over all fragments}
\label{sec:overview}

@@

The Worker Communication view consists of two charts: the aggregate communication
chart and the time series chart. The aggregate communication chart is a matrix (Figure~\ref{fig:network}) representing
a cluster of $N$ workers. Each cell $m,n$ in the matrix represents the total number of tuples sent
by worker $m$ to worker $n$ during the query execution. We chose to use the colors from color brewer\cite{harrower2003colorbrewer} and interpolated the colors between the steps using the \emph{CIE L*a*b*} interpolation and lightness correction \cite{chromajsaisch}. The colors work in grayscale and can be interpreted by color blind users.

The matrix chart representing aggregate communication between workers help uncover data skew
issues, as discussed in \S \ref{sec:eval}.

% We initially considered having a chord diagram\footnote{\url{http://bl.ocks.org/mbostock/4062006}} instead of a matrix
%for the aggregate communication chart. But feedback during the progress presentation session convinced us that
%a chord diagram would suffer from scalability problems. Adding an option for displaying a chord diagram remains
%part of future work.

@@

The current implementation focuses on scalability. We managed to some extent to
address the scalability issues raised by big query plans, large amount of
workers or long running queries. For example the \network view uses a
scalable heatmap to show the communication between all the workers in the
system. However, other techniques have to be further implemented to address
this issue. One example would be a more interactive query plan graph, which the
users can zoom. Also, using the ability to query the log data on the backend,
the \fragment view should only download and render what the user can see.

>>>>

We propose a visualization system to explore and understand query
execution and data movement in a distributed database management system
(DDBMS). Our tool provides insight into: (1) data flow between query segments
and between workers, (2) query execution and operator dependencies, (3) cluster
utilization, (4) network utilization. Our tool, \system, helps
developers understand and improve query execution, and provides insight into
common problems such as data skew or performance bottlenecks.

@@

Database query performance is often unpredictable, especially when the database is
distributed. This requires programmers or database developers to spend
significant amounts of time and effort isolating and debugging the root cause
of performance problems. These may include skewed data partitioning or run-time
issues, such as resource contention or adverse network conditions.

Our goal is to build a visualization tool designed to help a database developer understand the data flow and run-time
performance during query execution in a Distributed Database Management System (DDBMS). Using our tool, the
database programmer or developer could quickly discover run-time bottlenecks as well as problematic data skew issues,
saving them significant guesswork and programming effort to identify the appropriate logs to analyze. Our visualization tool,
\system, uses profiling logs from \emph{Myria} \cite{myria} as the prototype DDBMS. Myria is a distributed
database platform that provides \emph{big data management and analytics as a service} primarily for scientific applications.
Myria takes queries in high-level query languages, such as SQL, translating them into physical query plans and finally executing
them in parallel. Currently, Myria does not offer a visualization tool designed to show the physical query plan, job execution and network
communication between distributed workers. \system is designed to help the Myria programmer answer the following questions regarding
their query execution.

\begin{itemize}
\item \emph{How does the data flow through the system between operators?} 
The overall data flow in a DDBMS is represented by a \graph. It is important
for the programmer to understand the \graph in order to optimize it. To aid in this
understanding, we provide an interactive visualization of the \graph in \system.

%The physical
%query plan is composed of a directed graph of query \emph{fragments}. A query
%fragment is a subtree of the query execution plan that is separated from the
%rest of the plan by a network boundary. This means that the operators inside a
%query fragment can be executed by a single thread. Each fragment is scheduled
%to run on multiple workers, which means each worker thread executes the same set of
%operators inside the fragment, but on different data. Each graph link represents
%communication between workers, possibly over the network.  \system displays
%this \graph in an interactive fashion with clickable elements for fragment nodes
%and communication links.  The \graph also represents a high-level data flow in
%the DDBMS.

@@

The task of the back-end is to log when operators are called, when the call returns and when data is sent to another worker. Using this data and simple transformations, we can obtain the information necessary to create the visualizations to answer the questions in Section~\ref{sec:approach}.

There are two possible approaches that we explored to collect event logs that can be used to extract spans and other performance related data necessary for the visualization. First, events could be logged to files using a standard logging tool and then collect the files on one node using remote copy. Once the logs are on one machine, the logs can be parsed and the data can be transformed into the desired format.

We switched to a different approach as shown in Figure~\ref{fig:arch} where logs are written directly into the worker's database. When the front-end requests the data, the master executes a query that collects the data and streams the results directly to the requesting client. Logs are thus tuples in a relation and no longer separate in files and in a custom format. Despite being more reliable, faster and using the same logic that Myria already implements, we are able to write queries to aggregate, filter and transform the collected data.

For example, the data for how many workers are working on a certain fragment as described in Section~\ref{sec:fragments} is generated from the event log using a query to select only events for the root operators (the operators that have no parent in the same fragment) in each fragment and a custom map function that carries state. The state is an integer that is incremented when the operator is called on a worker and decremented when the operator returns.

@@

We highly suggest that developers use the second approach when porting \system to a new DDBMS.

\subsection{Front-end}

The Myria web front-end server is written in Python and runs on Google App Engine\footnote{\url{https://developers.google.com/appengine/}}. \system's user-interface (UI) is embedded into Myria's web front-end. We build \system's UI using D3\cite{2011-d3} to deliver an interactive visualization experience to the user. D3 is a JavaScript framework for data visualization on the web. We used visualization techniques such as \emph{focus+context}\cite{furnas1986generalized} that possible to implement conveniently using D3.

\system's front-end gets data from the Myria server as JSON and CSV files. JSON objects are used to describe the physical queryplan. CSV files contain logs on query execution at the fragment and operator level as well as data exchange between workers at different stages of the query execution.

@@

\todo{use the word spans when referring to events with duration (used in dapper)}

\subsubsection{Overview over all fragments}
\label{sec:fragments}

@@

The Worker Communication view consists of two charts: the aggregate communication
chart and the time series chart. The aggregate communication chart is a matrix representing
a cluster of $N$ workers. Each cell $m,n$ in the matrix represents the total number of tuples sent
by worker $m$ to worker $n$ during the query execution. Figure~\ref{fig:matrix} shows this chart along
with the selected link in the query plan DAG. Hovering the mouse over a cell brings up a tooltip
stating the total number of tuples exchanged between the source and destination workers. The intensity
of the cell color reflects the number of tuples exchanged, i.e., a larger number translates into a darker color
and a smaller number results in a lighter color. The color pallet used is grayscale and hence
immune to color-blindness issues. The color scale uses LAB interpolation.
The matrix chart representing aggregate communication
between workers help uncover data skew issues, as discussed in \S \ref{sec:eval}.

We initially considered having a chord diagram\footnote{\url{http://bl.ocks.org/mbostock/4062006}} instead of a matrix
for the aggregate communication chart. But feedback during the progress presentation session convinced us that
a chord diagram would suffer from scalability problems. Adding an option for displaying a chord diagram remains
part of future work.

@@

The current implementation focuses on scalability. We managed to some extent to
address the scalability issues raised by big query plans, large amount of
workers or long running queries. For example the \network view uses a
scalable heatmap to show the communication between all the workers in the
system. However, other techniques have to be further implemented to address
this issue. One example would be a more interactive query plan graph, which the
users can zoom. Another example, in the \fragment view that shows spans
for all events we should limit the data that is being downloaded and rendered
to what the user can see.
